Safety Analysis of the Molten Salt Fast Reactor Fuel Composition using Moltres

### 2019-06-17 paper/paper.tex

\begin{abstract}
%
    Abstract
\end{abstract}

@@

	While numerous computational tools exist for conventional nuclear reactors, 
    \glspl{MSR} present unique computational challenges that many fail to 
    address effectively. \glspl{MSR} differ
    profoundly from conventional 
    solid-fuelled reactors, particularly in their neutronics and 
    thermal-hydraulics behaviors.  New \gls{MSR} simulation tools must 
    capture strong coupling between neutronics and thermal-hydraulics 
    exhibited by \gls{DNP} movement as well as strong 
    Doppler and density feedback in the fuel salt.  This paper investigates 
    the impact of changing fuel composition on safety parameters in the 
    \gls{MSFR} concept using a new simulation tool for \glspl{MSR}, Moltres 
    \cite{lindsay_introduction_2018}.

@@

    The \gls{MSFR} model studied in this paper is a reference design for a 
    fast-spectrum \gls{MSR} developed under H2020
    \gls{SAMOFAR} project \cite{serp_molten_2014} (cite gerardin instead).
    The \gls{MSFR} boasts several safety and sustainability advantages over
    conventional reactors. Firstly, it can run on a closed thorium fuel cycle,
    which reduces actinide production and waste radiotoxity.
    The fast neutron spectrum improves $^{233}$U breeding from $^{232}$Th, an
    isotope that is about three times as abundant as uranium.
    Lastly, the \gls{MSFR} operates at near atmospheric pressure; this reduces
    the risk of containment structure failure and manufacturing costs.

@@

    We generated group constants for 
    each fuel composition using SERPENT \cite{leppanen_serpent_2015}, a 
    continuous-energy Monte Carlo code for numerous reactor physics 
    applications. For verification, we will compare group constants and
    reactivity coefficients with existing data 
    \cite{fiorina_analysis_2012,fiorina_investigation_2013} of safety 
    parameters for the three fuel compositions.  
    Using these group constants ($\chi$, $\nu\Sigma_f$, 
    $\Sigma_{g\rightarrow g'}$, $\frac{d\rho}{dT}$, etc.), Moltres then 
    solves for the flux and temperature based on the neutron diffusion 
    equation coupled with navier stokes thermal hydraulics. Transient 
    simulations will establish the spatial distribution of flux, 
    $\phi_g(\vec{r})$ and temperature, $T(\vec{r})$ during transients.
    These distributions will give insight into MSFR transient behavior,
    which will help us identify potential safety risks.
    These risks may warrant further study and possibly, changes to the
    fuel composition or the online fuel reprocessing scheme.

@@

	Lithium fluoride (LiF) is the major component of the fuel and blanket
	molten salts used for the \gls{MSFR}. Fissile and fertile isotopes are
	introduced
	into the mixture by mole fractions of 77.5\%LiF-22.5\%AcF$_4$, where
	AcF$_4$ represents actinide fluorides such as uranium and thorium
	fluorides.
	The \gls{MSFR} supports various fuel compositions. It can run on the same 
	$^{235}$U-$^{238}$U fuel used in most conventional LWRs. It can also run
	on a
	mixture of fresh and used uranium fuel containing reprocessed TRU
	isotopes \cite{fiorina_investigation_2013}. However, the main configuration
	of the \gls{MSFR} is a breeder reactor running on $^{233}$U-$^{232}$Th
	\cite{merle-lucotte_launching_2011}. Previous studies have reported
	breeding ratios of up to 1.1 on the \gls{MSFR} \cite{fiorina_molten_2013}.

@@

\begin{table}[t]
	\caption{Specifications of the \gls{MSFR} design \cite{serp_molten_2014}.}
	\begin{tabular}{ l l }
		\hline
		Parameter & Value \\
		\hline
		Thermal/Electric output [MW$_{\text{th}}$/MW$_{\text{e}}$] & 3000 /
		1500 
		\\
		Salt volume [m$^3$] & 18 \\
		Salt fraction in core & 0.5 \\
		Number of circulation loops & 16 \\
		Nominal flow rate [kg s$^{-1}$] & 18500  \\
		Nominal circulation time [s] & 4.0 \\
		Inlet/outlet temperature [K] & 923 / 1023 \\
		Blanket volume [m$^3$] & 7.3\\
		\hline
	\end{tabular}
	\label{table:msfr}
\end{table}

@@

	This section describes the overall modelling approach and the codes used
	in this paper. First, the \gls{BOL} and equilibrium compositions were
	obtained from SCALE/TRITON unit cell depletion calculations by Rykhlevskii
	et al. (cite Andrei) Their paper used a unit cell model for the modelling
	fuel cycle performance of the \gls{MSFR} and three other fast-spectrum
	\glspl{MSR}. The neutron spectrum and depletion calculations on the unit
	cell SCALE/TRITON models were verified against full core SERPENT models and
	showed very good agreement. The equilibrium composition was selected at
	approximately 43 years after start-up because the TRU vector change
	between each depletion time-step from this moment onwards is less than 3\%.
	(cite Andrei) For the \gls{BOL} composition, it is the fuel composition at
	300 days after start-up. Decay isotopes, whose neutron cross-section data
	are not found in the
	JEFF-3.1.2 nuclear data library, were omitted.

	Neutronics calculations were performed on SERPENT using these fuel
	compositions and a truncated, reference \gls{MSFR} core model
	\cite{pettersen_coupled_2016} to generate six-group neutronics data for
	Moltres.
	The energy boundaries for the six neutron groups are adopted from previous
	\gls{MSFR} studies and shown in Table \ref{table:bound}.
	As for the \gls{DNP} groups, there are eight groups as defined by the
	JEFF-3.1.2 nuclear data library used for neutron cross-section data.
	The reactor model is a 2D axisymmetric model as shown in
	Figure \ref{fig:reference}. This geometry focuses on the active central
	region of the core. Basing the geometry on the reference model facilitates
	comparisons with previous MSFR studies by various authors
	\cite{fiorina_modelling_2014} \cite{pettersen_coupled_2016} on the
	reference model for code-to-code verification purposes.
	The 2D model is extended into a 3D model in SERPENT by rotating it
	around the central axis to form concentric cylinders.

@@

	Finally, with the group constant data from SERPENT, Moltres
	calculated the resulting coupled neutron spectrum and temperature
	distribution for a given velocity profile in the core, from user-specified
	initial conditions towards steady state, or from steady state in the case
	of an accident scenario. The mesh geometry for Moltres is
	further simplified from the SERPENT model by the omission of the absorber,
	heat exchanger, and reflector regions. This is justified as they have
	negligible contribution towards the neutronics. Instead, a vacuum
	boundary condition was imposed on the outer surface of the blanket region,
	along with 923 K Dirichlet boundary conditions for temperature.
	The 2 cm thick blanket tank structural material was also omitted as its
	inclusion resulted in an extra fine mesh which drastically
	increased computational time. (Insert Moltres mesh model)

	Moltres has a loop app functionality to account for the flow and decay of
	\glspl{DNP} outside the active core region. For this paper, the two ends of
	a separate, 188 cm long 1-D outer
	loop region were ``attached" to the inlet and outlet of active core region
	model, shown in Figure *, to form a closed coolant loop. This was achieved
	using \gls{DNP} boundary
	conditions at the inlets and outlets of the core and outer loop geometry,
	determined by the inflow and outflow values of each \gls{DNP} group at
	each time-step for each boundary.
	\gls{DNP} in the outer loop are also governed by the same \gls{DNP}
	equations to allow for flow, production, and decay.

@@

\begin{figure}[t] 
	\centering
	\includegraphics[width=0.46\textwidth]{./figures/reference}
	\captionsetup{justification=centering}
	\caption{Cross-section of the 2D axisymmetric model used for the
	simulation. Derived from the 2D axisymmetric reference model of the MSFR
	\cite{pettersen_coupled_2016}. The figure is not drawn to scale.}
	\label{fig:reference}
\end{figure}

@@

@@

	Total power as defined by Eq. \ref{eq4} is approximately 2 GW. This
	is 1 GW less than the rated 3 GW of the \gls{MSFR}. This may be mainly due
	to the erroneous uniform velocity profile imposed in the core. From
	fluid dynamics, velocity along the central axis of the core should be
	higher than the velocity at the periphral areas near the blanket tank.
	The higher velocity at the center results in higher advective heat transfer
	away from the center of the core and cools the active region. This should
	result in an increase in neutron flux from the strongly negative
	temperature reactivity feedback, and thus higher heat generation. The
	truncation of the top and bottom fuel salt region is also a factor for
	the lower than expected heat production.

	Comparing the three different fuel compositions, the \gls{MSFR} operates at
	the highest temperatures with the start-up fuel composition, followed by
	the \gls{BOL}, and the equilibrium compositions. This may be due to the
	generation of $^{233}$Pa as an intermediate for $^{233}$U breeding as
	mentioned by Rykhlevskii et al. (cite Andrei) in their fuel cycle analysis
	of the \gls{MSFR}.

@@

	The loss of heat sink is expected to cause an initial temperature spike,
	followed by a decrease in neutron flux and heat generation due to the
	strongly negative temperature reactivity feedback. Eventually, the
	\gls{MSFR} should settle on a new steady state with a higher average
	temperature than the operating temperature. If the high temperature is
	sustained for a long period of time, it could lead to severe damage
	in the reactor, pipes, or other instruments.

@@

	Summarise results

	Safe

	Further development of the models is necessary to improve the accuracy
	and reliability of Moltres as a coupled neutronics/thermal-hydraulics code.
	Further work includes using an arbitrarily defined velocity profile, such
	as a parabolic profile. A better option would be to

>>>>

\begin{abstract}
%
    The Molten
    Salt Fast Reactor (MSFR) has garnered great interest for its inherent
    safety and sustainbility features. The MSFR can adopt a closed thorium fuel
    cycle for
    sustainable operation through the breeding of $^{233}$U from $^{232}$Th.
    The fuel composition changes significantly over the course of its lifespan.
    In this study, we investigated the steady state and transient behavior of
    the MSFR using Moltres, a coupled neutronics/thermal-hydraulics code
    developed within the Multiphysics Object Oriented Simulation Environment
    (MOOSE) framework. Three different fuel compositions, start-up,
    beginning-of-life (BOL), and equilibrium, were examined for potentially
    dangerous
    core temperature excursions during a loss of secondary cooling accident
    (LOSCA). The steady state peak neutron flux showed good agreement
    with existing MSFR data, while the temperature distribution and total power
    showed discrepancies which can be attributed to known sources of error.
    For the transient behavior under the LOSCA
    scenario, the time delay towards the new steady state core temperature is
    also in good agreement with existing MSFR data. Moltres underestimated the
    temperature rise by a factor of ten relative to other published data, due
    to the same sources of
    error for the steady state results. While a MSFR loaded with
    start-up fuel composition operates at a higher temperature than with the
    other two fuel compositions, all three cases were shown to be inherently
    safe due to the strong negative temperature feedback.
\end{abstract}

@@

	While numerous computational tools exist for conventional nuclear reactors, 
    \glspl{MSR} present unique computational challenges that many fail to 
    address effectively. \glspl{MSR} differ
    profoundly from conventional 
    solid-fuelled reactors, particularly in their neutronics and 
    thermal-hydraulics behaviors.  New \gls{MSR} simulation tools must 
    capture strong coupling between neutronics and thermal-hydraulics 
    exhibited by \gls{DNP} movement as well as strong 
    Doppler and density feedback in the fuel salt.  This paper investigates 
    the impact of changing fuel composition on fuel salt temperatures in the 
    \gls{MSFR} concept using a new simulation tool for \glspl{MSR}, Moltres 
    \cite{lindsay_introduction_2018}.

@@

    The \gls{MSFR} model studied in this paper is a reference design for a 
    fast-spectrum \gls{MSR} developed under H2020
    \gls{SAMOFAR} project \cite{gerardin_design_nodate}.
    The \gls{MSFR} boasts several safety and sustainability advantages over
    conventional reactors. Firstly, it can run on a closed thorium fuel cycle,
    which reduces actinide production and waste radiotoxity.
    The fast neutron spectrum improves $^{233}$U breeding from $^{232}$Th, an
    isotope that is about three times as abundant as uranium.
    Lastly, the \gls{MSFR} operates at near atmospheric pressure; this reduces
    the risk of containment structure failure and manufacturing costs.

@@

    We generated group constants for 
    each fuel composition using SERPENT \cite{leppanen_serpent_2015}, a 
    continuous-energy Monte Carlo code for numerous reactor physics 
    applications.
    Using these group constants, Moltres then 
    solves for the flux and temperature based on the neutron diffusion 
    equation coupled with temperature advection due to coolant flow. Transient 
    simulations will establish core fuel temperatures during transients.
    These distributions will give insight into MSFR transient behavior,
    which will help us identify potential safety risks.

@@

	Lithium fluoride (LiF) is the major component of the fuel and blanket
	molten salts used for the \gls{MSFR}. Fissile and fertile isotopes are
	introduced
	into the mixture by mole fractions of 77.5\%LiF-22.5\%AcF$_4$, where
	AcF$_4$ represents actinide fluorides such as uranium and thorium
	fluorides. Table \ref{table:fuelsalt} shows the mole fraction of the
	molecular constituents in the \gls{MSFR} fuel salt.
	The \gls{MSFR} supports various fuel compositions; it can run on the same 
	$^{235}$U-$^{238}$U fuel used in most conventional LWRs. It can also run
	on a
	mixture of fresh and used uranium fuel containing reprocessed TRU
	isotopes \cite{fiorina_investigation_2013}. However, the main configuration
	of the \gls{MSFR} is a breeder reactor running on $^{233}$U-$^{232}$Th
	\cite{merle-lucotte_launching_2011}. Previous studies have reported
	breeding ratios of up to 1.1 on the \gls{MSFR} \cite{fiorina_molten_2013}.

@@

\begin{table}[t]
	\caption{Specifications of the \gls{MSFR} design \cite{serp_molten_2014}.}
	\begin{tabular}{ l l }
		\hline
		Parameter & Value \\
		\hline
		Thermal/Electric output [MW$_{\text{th}}$/MW$_{\text{e}}$] & 3000 /
		1500 
		\\
		Salt volume [m$^3$] & 18 \\
		Salt fraction in core & 0.5 \\
		Number of circulation loops & 16 \\
		Nominal flow rate [kg s$^{-1}$] & 18500  \\
		Nominal circulation time [s] & 4.0 \\
		Inlet/outlet temperature [K] & 923 / 1023 \\
		Blanket volume [m$^3$] & 7.3\\
		\hline
	\end{tabular}
	\label{table:msfr}
\end{table}
%
\begin{table}[t]
\centering
\begin{tabular}{rr}
\hline
{Isotope} & {Mole fraction [\%]}\\
\hline
$^7$Li & 29.0\\
$^{19}$F & 62.6\\
$^{232}$Th & 7.40\\
$^{233}$U & 1.00\\
\hline
\end{tabular}
\captionsetup{justification=centering}
\caption{Fuel salt composition used for the Serpent simulation.}
\label{table:fuelsalt}
\end{table}

@@

	This section describes the overall modelling approach and the codes used
	in this paper. First, the \gls{BOL} and equilibrium compositions were
	obtained from SCALE/TRITON unit cell depletion calculations by Rykhlevskii
	et al. \cite{rykhlevskii_fuel_2019} Their paper used a unit cell model for
	the modelling
	fuel cycle performance of the \gls{MSFR} and three other fast-spectrum
	\glspl{MSR}. The neutron spectrum and depletion calculations on the unit
	cell SCALE/TRITON models were verified against full core SERPENT models and
	showed very good agreement. We selected the equilibrium composition at
	approximately 43 years after start-up because the TRU vector change
	between each depletion time-step from this moment onwards is less than 3\%
	\cite{rykhlevskii_fuel_2019}. For the \gls{BOL} composition, it is the 
	resulting fuel composition at
	300 days after start-up. Decay isotopes, whose neutron cross-section data
	are not found in the
	JEFF-3.1.2 nuclear data library, were omitted.

	We performed neutronics calculations on SERPENT using these fuel
	compositions and a truncated, reference \gls{MSFR} core model
	\cite{pettersen_coupled_2016} to generate six-group neutronics data for
	Moltres.
	The energy boundaries for the six neutron groups were adopted from previous
	\gls{MSFR} studies and shown in Table \ref{table:bound}.
	As for the \gls{DNP} groups, there are eight groups as defined by the
	JEFF-3.1.2 nuclear data library used for neutron cross-section data.
	The reactor model is a 2D axisymmetric model as shown in
	Figure \ref{fig:reference}. This geometry focuses on the active central
	region of the core. Basing the geometry on the reference model facilitates
	comparisons with previous MSFR studies by various authors
	\cite{fiorina_modelling_2014} \cite{pettersen_coupled_2016} on the
	reference model for code-to-code verification purposes.
	The 2D model is extended into a 3D model in SERPENT by rotating it
	around the central axis to form concentric cylinders.

@@

	Finally, with the group constant data from SERPENT, Moltres
	calculated the resulting coupled neutron spectrum and temperature
	distribution for a given velocity profile in the core, from user-specified
	initial conditions towards steady state, or from steady state in the case
	of an accident scenario. We firther simplified the mesh geometry for
	Moltres from the SERPENT model by omitting the absorber,
	heat exchanger, and reflector regions. This is justified as they have
	negligible contribution towards the neutronics. Instead, a vacuum
	boundary condition was imposed on the outer surface of the blanket region,
	along with 923 K Dirichlet boundary conditions for temperature.
	The 2 cm thick blanket tank structural material was also omitted as its
	inclusion resulted in an extra fine mesh which drastically
	increased computational time. Figure \ref{fig:mesh} shows the Moltres
	\gls{MSFR} mesh.

	Moltres has a loop app functionality to account for the flow and decay of
	\glspl{DNP} outside the active core region. For this paper, the two ends of
	a separate, 188 cm long 1-D outer
	loop region were ``attached" to the inlet and outlet of active core region
	model, shown in Figure *, to form a closed coolant loop. We achieved this
	by using \gls{DNP} boundary
	conditions at the inlets and outlets of the core and outer loop geometry,
	determined by the inflow and outflow values of each \gls{DNP} group at
	each time-step for each boundary.
	The \gls{DNP} equations also govern the \gls{DNP} behavior in the outer
	loop to allow for flow, production, and decay.

@@

@@

\begin{figure}[t] 
	\centering
	\includegraphics[width=0.47\textwidth]{./figures/reference}
	\captionsetup{justification=centering}
	\caption{Cross-section of the 2D axisymmetric model used in SERPENT.
	Derived from the 2D axisymmetric reference model of the MSFR
	\cite{pettersen_coupled_2016}. Figure is not drawn to scale.}
	\label{fig:reference}
\end{figure} 
%
\begin{figure}[t] 
	\centering
	\includegraphics[width=0.45\textwidth]{./figures/mesh}
	\captionsetup{justification=centering}
	\caption{Mesh of the 2D axisymmetric model used in Moltres.
	The grey and red regions represent the fuel and blanket salt respectively.}
	\label{fig:mesh}
\end{figure}

@@

	Total power as defined by Eq. \ref{eq4} is approximately 2 GW. This
	is 1 GW less than the rated 3 GW of the \gls{MSFR}. This may be mainly due
	to the erroneous uniform velocity profile imposed in the core. From
	fluid dynamics, velocity along the central axis of the core should be
	higher than the velocity at the periphral areas near the blanket tank.
	The higher velocity at the center results in higher advective heat transfer
	away from the center of the core and cools the active region. This should
	result in an increase in neutron flux from the strong negative
	temperature reactivity feedback, and thus higher heat generation. The
	truncation of the top and bottom fuel salt region is also a factor for
	the lower than expected heat production.

	Comparing the three different fuel compositions, the \gls{MSFR} operates at
	the highest temperatures with the start-up fuel composition, followed by
	the \gls{BOL}, and the equilibrium compositions. This may be due to the
	generation of $^{233}$Pa as an intermediate for $^{233}$U breeding as
	mentioned by Rykhlevskii et al. \cite{rykhlevskii_fuel_2019} in their fuel cycle analysis
	of the \gls{MSFR}; $^{233}$Pa is a significant neutron poison.

@@

	The loss of heat sink is expected to cause an initial temperature spike,
	followed by a decrease in neutron flux and heat generation due to the
	strong negative temperature reactivity feedback. Eventually, the
	\gls{MSFR} should settle on a new steady state with a higher average
	temperature than the operating temperature. If the high temperature is
	sustained for a long period of time, it could lead to severe damage
	in the reactor, pipes, or other instruments.

@@

	In this study, we investigated the steady state, and transient behavior of
	the \gls{MSFR} during a \gls{LOSCA}
	using the coupled neutronics/thermal-hydraulics code Moltres. Moltres
	boasts relatively fast computational times due to a combination of
	parallelization,
	adaptive coarse meshing scheme and adaptive time-stepping. The total
	neutron flux at steady state agrees very closely with published results by
	Fiorina et al. \cite{fiorina_modelling_2014} While the temperature
	distribution and total power have
	large discrepancies, we have accounted for the main sources of error:
	the erroneous uniform velocity profile and the reactor model truncation. We
	also observed that the \gls{MSFR} operates at a higher temperature with the
	start-up fuel composition than with the \gls{BOL} and equilibrium
	compositions, due to the presence of $^{233}$Pa.

	During a \gls{LOSCA}, the \gls{MSFR} core fuel salt temperatures rise by
	approximately 10 K after a time delay of 30-40 s. While the time delay is
	similar to published results, the temperature rise is relatively low
	compared to the 100 K rise reported by Fiorina et al.
	\cite{fiorina_modelling_2014} This discrepancy is also attributed to the
	same sources of error for the steady state results. Nonetheless, a
	temperature rise of
	this magnitude is relatively benign to reactor components, and makes
	a strong safety case for \glspl{MSR} due to their strong negative
	temperature feedback.

	Further development of the models is necessary to improve the accuracy
	and reliability of Moltres as a coupled neutronics/thermal-hydraulics code.
	Further work includes using an arbitrarily defined velocity profile, such
	as a parabolic profile. A better option would be the development of a
	Navier-Stokes module to generate an accurate velocity profile.

	The heat exchanger functionality in Moltres can also be improved beyond
	the current implementation of a fixed heat sink. The relevant heat
	transfer parameters between the primary and secondary loop in the heat
	exchanger would be necessary for a more complicated heat exchanger system.
